{RandomRationalPoints} package for \emph{Macaulay2}

### 2020-11-10 FastLinAlg/Paper/RandomRationalPoints.tex
Examples 4.1 and 4.2 added. (Z.J.)
Various changes throughout.

\documentclass[11pt]{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,fullpage    }
%\usepackage{mathtools}
\RequirePackage[dvipsnames,usenames, table]{xcolor}
\usepackage{hyperref,cleveref}
\hypersetup{
bookmarks,
bookmarksdepth=3,
bookmarksopen,
bookmarksnumbered,
pdfstartview=FitH,
colorlinks,backref,hyperindex,
linkcolor=Sepia,
anchorcolor=BurntOrange,
citecolor=MidnightBlue,
citecolor=OliveGreen,
filecolor=BlueViolet,
menucolor=Yellow,
urlcolor=OliveGreen
}
\usepackage{xypic}
\usepackage{alltt}
\usepackage{multicol}  
%\usepackage{etex}
\usepackage{xspace}
\usepackage{rotating}
\interfootnotelinepenalty=100000
\theoremstyle{definition}
\newtheorem{example}{Example}[section]

@@

 While there is an existing package called {\tt RandomPoints}, which we took inspiration from, which aims to find all the rational points of a variety, our aim here is to find one point quickly.
 We provide some functions that apply different strategies to generate random rational points for the given variety. We also provide functions that will expedite the process of determining properties of the singular locus of $X$ when combined with the functions provided in the package {\tt FastLinAlg}.

@@

All polynomial rings that we will consider will be over finite fields.

We start {\tt randomCoordinateChange}, a helper function that also might be useful to others.

%	\item {\tt Codimension}: It takes input an integer. Various functions need to know the codimension/height of the scheme/ideal it is working with.  Setting this to be an integer will tell the function not to compute the codimension and to use this value instead.  The default value is {\tt null}, in which case the function will compute the codimension.  
%%%
%%Codimension is not a function, it is an option.
%%%
\subsection*{\tt randomCoordinateChange}: It takes a polynomial ring as an input and outputs the coordinate change map. 
	i.e. given a polynomial ring, this will produce a linear automorphism of the ring.  This function checks whether the map is an isomorphism by computing the Jacobian.

	In some applications, a full change of coordinates is not desired, as it might cause code to run very slowly.  A binomial change of coordinates might be preferred, or we might only replace some coordinates by monomials.  
	Refinements of {\tt randomCoordinateChange} are controlled with the following options.

	\begin{itemize}
		\item {\tt MaxCoordinatesToReplace}: the user can specify that only a specified number of coordinates should be non-monomial (assuming {\tt Homogeneous => true}).  
		This option is passed to {\tt randomCoordinateChange} by other functions that call it. 		
		\item {\tt Homogeneous}:  Setting {\tt Homogeneous => false} will cause degree zero terms to be added to modified coordinates (including monomial coordinates).
		\item {\tt Replacement}: setting {\tt Replacement => Full} will mean that coordinates are replaced by a general degree $1$ form.  If {\tt Replacement => Binomial}, the coordinates are only changed to binomials, which can be much faster for certain applications. If {\tt Homogeneous => false}, then there will be constant terms, and we view $mx + b$ as a binomial.
	\end{itemize}

@@

\begin{enumerate}
\item {\tt genericProjection} 
finds a random (somewhat) generic projection of the ring or ideal.
The typical usages are as follows: 
\vspace{1em}

@@

This gives the projection map from $\mathbb{A}^N \mapsto\mathbb{A}^{N-n}$ and the defining ideal of the projection of $V(I)$. If no integer $n$ is provided, then drops one dimension, in other words it treats $n = 1$.

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}
i1 : R=ZZ/5[x,y,z,w];

@@

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}
i1 : R=ZZ/5[x,y,z,w];

@@

\vspace{1em} 
\item {\tt projectionToHypersurface} creates a projection to a hypersurface. The typical usages are as follows:

It differs from {\tt genericProjection(codim I - 1, I)} as it only tries to find a hypersurface equation that vanishes along the projection, instead of finding one that vanishes exactly at the projection.  This can be faster, and can be useful for finding points.

If we already know the codimension is {\tt c}, we can set {\tt Codimension=>c} so the function does not compute it. The user can also specify the same options as before to be passed on to the {\tt randomCoordinateChange} function.

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}
i2 : R=ZZ/5[x,y,z,w];

@@

\end{enumerate}
\section{randomPoints}\label{randomPoints}
{\tt randomPoints} is a function to find rational or geometric points in a variety. The typical usages are as follows:

@@

\subsection{Additional Options}\label{strategydetails}

The user may also choose to provide some additional information depending on the context which may help in faster computations.

\begin{enumerate}
	\vspace{1em}

\item {\tt Strategy => $\cdot$} where the $\cdot$ can be \textit{Default, BruteForce, LinearIntersection, GenericProjection and  HybridProjectionIntersection}.

@@

	\item {\tt HybridProjectionIntersection} does a generic projection, followed by a linear intersection. Notice that speed, or success, varies depending on the strategy.

\end{itemize}

\begin{example}
	content...
\end{example}

\item {\tt ProjectionAttempts => ZZ}:

@@

\item {\tt MaxCoordinatesToReplace => ZZ} (see \Cref{helper})

\vspace{1em}
\item {\tt Codimension => ZZ} (see \Cref{helper})

\vspace{1em}
\item {\tt ExtendField => Boolean}:

@@

\vspace{1em}
\item {\tt IntersectionAttempts => ZZ}:

@@

\vspace{1em}
\item{\tt PointCheckAttempts => ZZ}:

@@

 \begin{example}
 	content...
 \end{example}

\vspace{1em}
\item {\tt MaxCoordinatesToTrivialize}:

@@

\vspace{1em}
\item {\tt NumThreadsToUse => ZZ}:

When calling {\tt randomPoints}, and functions that call it, with a {\tt BruteForce} strategy, this denotes the number of threads to use in brute force point checking.
\end{enumerate}
%Outputs
%:List
%a list of points in the variety with possible repetitions.
%Description
%Text  
%Gives at most $n$ many point in a variety $V(I)$. 
%Example
%R = ZZ/5[t_1..t_3];
%I = ideal(t_1,t_2+t_3);
%randomPoints(3, I)
%randomPoints(4, I, Strategy => Default)
%randomPoints(4, I, Strategy => GenericProjection)
%randomPoints(4, I, Strategy => LinearIntersection)

@@

\begin{itemize}
	\item {\tt Strategy => Symbol} to specify which strategy to use when calling {\tt randomPoints} (see \Cref{strategydetails}).

	\item {\tt Verbose => Boolean}
		 Set the option {\tt Verbose => true} to turn on verbose output.  This may be useful in debugging or in determining why an computation is running slowly.

	\item {\tt Homogeneous => Boolean} (see \Cref{helper})

	\item {\tt MinorPointAttempts => ZZ} :
	how many points to check the rank of the matrix at
\end{itemize}

@@

%\vspace{1em}
As again, we can have the following additional options:

\vspace{1em}
-- {\tt Strategy => Symbol}: specify which strategy to use when calling {\tt randomPoints}

-- {\tt Homogeneous => Boolean}: controls if the computations are homogeneous (in calls to {\tt randomPoints})

-- {\tt Verbose => Boolean}: turns on or off verbose output (see \Cref{findANonZeroMinor})

-- {\tt MinorPointAttempts => ZZ}: the maximum number of points to check the rank of the matrix at

>>>>

\documentclass[11pt]{amsart}
\usepackage{calc,amssymb,amsthm,amsmath,fullpage    }
%\usepackage{mathtools}
\RequirePackage[dvipsnames,usenames, table]{xcolor}
\usepackage{hyperref}
\usepackage[capitalize]{cleveref}
\hypersetup{
bookmarks,
bookmarksdepth=3,
bookmarksopen,
bookmarksnumbered,
pdfstartview=FitH,
colorlinks,backref,hyperindex,
linkcolor=Sepia,
anchorcolor=BurntOrange,
citecolor=MidnightBlue,
citecolor=OliveGreen,
filecolor=BlueViolet,
menucolor=Yellow,
urlcolor=OliveGreen
}
\usepackage{xypic}
\usepackage{alltt}
\usepackage{multicol}  
%\usepackage{etex}
\usepackage{xspace}
\usepackage{rotating}
\interfootnotelinepenalty=100000
\theoremstyle{definition}
\newtheorem{example}{Example}[section]

@@

 There is an existing package called {\tt RandomPoints}, which we took inspiration from, which aims to find \emph{all} the rational points of a variety, our aim here is to find one point quickly, even if it is not rational.  
 We provide some functions that apply different strategies to generate random rational points for the given variety. We also provide functions that will expedite the process of determining properties of the singular locus of $X$ when combined with the functions provided in the package {\tt FastLinAlg}.

@@

%We start {\tt randomCoordinateChange}, a helper function that also might be useful to others.

%	\item {\tt Codimension}: It takes input an integer. Various functions need to know the codimension/height of the scheme/ideal it is working with.  Setting this to be an integer will tell the function not to compute the codimension and to use this value instead.  The default value is {\tt null}, in which case the function will compute the codimension.  
%%%
%%Codimension is not a function, it is an option.
%%%
\subsection*{\tt randomCoordinateChange} This function takes a polynomial ring as an input and outputs the coordinate change map.  
	i.e. given a polynomial ring, this will produce a linear automorphism of the ring.  This function checks whether the map is an isomorphism by computing the Jacobian.

	In some applications, a full change of coordinates is not desired, as it might cause code to run very slowly.  A binomial change of coordinates might be preferred, or we might only replace some monomials by other monomials.  
	Refinements of {\tt randomCoordinateChange} are controlled with the following options.

	\begin{itemize}
		\item {\tt Replacement}: setting {\tt Replacement => Full} will mean that coordinates are replaced by a general degree $1$ form.  If {\tt Replacement => Binomial}, the coordinates are only changed to binomials, which can be much faster for certain applications. If {\tt Homogeneous => false}, then there will be constant terms, and we view $mx + b$ as a binomial.
		\item {\tt MaxCoordinatesToReplace}: the user can specify that only a specified number of coordinates should be non-monomial (assuming {\tt Homogeneous => true}).  
		This option is passed to {\tt randomCoordinateChange} by other functions that call it. 		
		\item {\tt Homogeneous}:  Setting {\tt Homogeneous => false} will cause degree zero terms to be added to modified coordinates (including monomial coordinates).		
	\end{itemize}

@@

%\begin{enumerate}
\subsection{genericProjection} 
This function finds a random (somewhat depending on options) generic projection of the ring or ideal.
The typical usages are as follows: 
\vspace{1em}

@@

This gives the projection map from $\mathbb{A}^N \mapsto\mathbb{A}^{N-n}$ and the defining ideal of the projection of $V(I)$. If no integer $n$ is provided then it acts as if $n = 1$.

\begin{example}	
	{{\small\color{blue}
	\begin{verbatim}

i1 : R=ZZ/5[x,y,z,w];

@@

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}

i1 : R=ZZ/5[x,y,z,w];

@@

The user can also specify the same options as before to be passed on to the {\tt randomCoordinateChange} function.

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}

i2 : R=ZZ/5[x,y,z,w];

@@

%\vspace{1em} 
\subsection{\tt projectionToHypersurface} This function creates a projection to a hypersurface. The typical usages are as follows:

It differs from {\tt genericProjection(codim I - 1, I)} as it only tries to find a hypersurface equation that vanishes along the projection, instead of finding one that vanishes exactly at the projection.  This can be faster, and can be useful for finding points.   If we already know the codimension is {\tt c}, we can set {\tt Codimension=>c} so the function does not compute it.

%\end{enumerate}
\section{randomPoints}\label{randomPoints}
{\tt randomPoints} is a function to find rational or geometric points in a variety. The typical usages are as follows:

@@

\subsection{Options}\label{strategydetails}

The user may also choose to provide some additional information depending on the context which may help in faster computations, or whether a point is found at all.

\begin{description}		
	\item[\tt Strategy => $\bullet$]   Here the $\bullet$ can be {\tt Default, BruteForce, LinearIntersection, GenericProjection} or {\tt HybridProjectionIntersection}.

@@

		\item {\tt HybridProjectionIntersection} does a generic projection, followed by a linear intersection. 		
	\end{itemize}

	Notice that speed, or success, varies depending on the strategy.

	\begin{example}\label{BruteForce}
		{{\small\color{blue}
		\begin{verbatim}

			i2 : R = ZZ/101[x,y,z];

			i3 : J = ideal(x^3+y^2+1,z^3-x^2-y^2+2);

			i4 : randomPoints(J)
				-- used 0.0181549 seconds

			o4 = {{32, 37, -30}}

			o4 : List

			i5 : time randomPoints(J,Strategy=>BruteForce)
				-- used 0.0146932 seconds

			o5 = {}

			o5 : List

			i6 : randomPoints(J,Strategy=>GenericProjection)
				-- used 0.0146932 seconds

			o6 = {{2, -30, -8}}

			o6 : List

	\end{verbatim}
	}}
	\end{example}

	\item[\tt ProjectionAttempts => ZZ]

@@

	\vspace{1em}
	\item[\tt MaxCoordinatesToReplace => ZZ] (see \Cref{helper})
	\vspace{1em}
	\item[\tt Codimension => ZZ] (see \Cref{helper})

	\vspace{1em}
	\item[\tt ExtendField => Boolean]

@@

	\vspace{1em}
	\item[\tt IntersectionAttempts => ZZ]

@@

	\vspace{1em}
	\item[\tt PointCheckAttempts => ZZ]

@@

	\begin{example}
	We re-compute \cref{BruteForce} this time specifying more attempts.
		{{\small\color{blue}
		\begin{verbatim}
	i1: R = ZZ/101[x,y,z];

	o1 : PolynomialRing

	i2 : J = ideal "x3+y2+1,z3-x2-y2+2";

	o2 : Ideal of R

	i3 : randomPoints(J,Strategy=>BruteForce,PointCheckAttempts=>20000)
		-- used 0.84679 seconds

	o3 = {{-39, -43, 28}}

	o3 : List
	\end{verbatim}
	}}
	\end{example}

	\vspace{1em}
	\item[\tt MaxCoordinatesToTrivialize]

@@

	\vspace{1em}
	\item[\tt NumThreadsToUse => ZZ]

	When calling {\tt randomPoints}, and functions that call it, with a {\tt BruteForce} strategy, this denotes the number of threads to use in brute force point checking.
\end{description}
%Outputs
%:List
%a list of points in the variety with possible repetitions.
%Description
%Text  
%Gives at most $n$ many point in a variety $V(I)$. 
%Example
%R = ZZ/5[t_1..t_3];
%I = ideal(t_1,t_2+t_3);
%randomPoints(3, I)
%randomPoints(4, I, Strategy => Default)
%randomPoints(4, I, Strategy => GenericProjection)
%randomPoints(4, I, Strategy => LinearIntersection)

\subsection{Comments on performance}

	When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly.
	\begin{example}
		{{\small\color{blue}
		\begin{verbatim}

			i2 : R = ZZ/7[x_1..x_10]

			o2 = R

			o2 : PolynomialRing

			i3 : I = ideal(random(2, R), random(3,R));

			i4 : time randomPoints(I,Strategy=>BruteForce, PointCheckAttempts=>20000)
				 -- used 0.0102075 seconds

			o4 = {{-2, -1, 3, -2, -2, 2, 0, -2, -2, -2}}

			o4 : List

			i5 : time randomPoints(I,Strategy=>Default)
				 -- used 0.085741 seconds

			o5 = {{1, -2, 2, 2, -2, 0, 0, -1, -1, 3}}

			o5 : List

			i6 : S = ZZ/211[x_1..x_10]

			o6 = S

			o6 : PolynomialRing

			i7 : J = ideal(random(2, S), random(3, S));

			o7 : Ideal of S

			i8 : time randomPoints(J,Strategy=>BruteForce, PointCheckAttempts=>2000000)
				-- used 25.3479 seconds

			o8 = {{-36, 26, -63, -7, -48, 78, -44, -22, 105, 39}}

			i9 : time randomPoints(J,Strategy=>Default)
			-- used 0.102918 seconds

			o9 = {{0, -42, -53, -33, -65, -27, -105, -37, 99, 64}}

			o9 : List

			i10 : time randomPoints(J,Strategy=>LinearIntersection)
			-- used 0.0169781 seconds

			o10 = {{-51, -11, -34, -33, -65, 19, -86, 50, -44, -97}}

			o10 : List

			i11 : time randomPoints(J,Strategy=>GenericProjection)
			-- used 0.364902 seconds

			o11 = {{31, -15, 53, 41, 87, 9, -21, -43, -12, 85}}

			o11 : List

			i12 : time randomPoints(J,Strategy=>LinearIntersection, ExtendField => true)
     		-- used 0.0612306 seconds

			               5      4      3      2                                      5  
			o12 = {{54, 14a  - 48a  - 56a  + 56a  - 13a + 48, -101, 56, -42, -27, - 26a  -
				-------------------------------------------------------------------------
				   4      3      2
				43a  - 70a  - 96a  + 50a - 103, -29, 0, -37}}

	\end{verbatim}
	}}
	\end{example}

@@

\begin{description}
	\item[\tt Strategy => Symbol] to specify which strategy to use when calling {\tt randomPoints} (see \Cref{strategydetails}).

	\item[\tt Verbose => Boolean]
		 Set the option {\tt Verbose => true} to turn on verbose output.  This may be useful in debugging or in determining why an computation is running slowly.

	\item[\tt Homogeneous => Boolean] (see \Cref{helper})

	\item[\tt MinorPointAttempts => ZZ] 
		This controls how many points at which to check the rank of the matrix.
\end{description}

@@

%\vspace{1em}
As again, we can have the following additional options:
\begin{description}
\item[\tt Strategy => Symbol] This specifies which strategy to use when calling {\tt randomPoints}.
\item[\tt Homogeneous => Boolean] This is passed in calls to {\tt randomPoints}.
\item[\tt Verbose => Boolean] This turns on or off verbose output (see \Cref{findANonZeroMinor})
\item[\tt MinorPointAttempts => ZZ] This controls how many points at which to check the rank of the matrix.
\end{description}
